Studying measurement invariance in factor analysis models via stochastic processes

### 2011-02-11 MIpaper/par_instable.tex
Revised manuscript (except simulation section)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@45 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

To employ the LR test for assessing measurement invariance, model
parameters are estimated separately for a certain number of subgroups
of the data (with some parameters potentially restricted to be equal across
subgroups).\footnote{For ease of exposition we present
the results for the case where there are no such restrictions, although it is
straightforward to extend all methods to the more general case.}
Then, the sum of maximized likelihoods 
from the subgroups are compared with the original maximized full-sample likelihood
in a $\chi^2$~test. More specifically, for the special case of two subgroups,
the alternative $H_1^*$ from (\ref{eq:h1*}) with fixed and prespecified $\nu$ is adopted
and the null hypothesis $H_0$ from (\ref{eq:h0}) reduces to ${\bm \theta}^{(A)} = {\bm \theta}^{(B)}$.
The parameter estimated $\hat {\bm \theta}^{(A)}$ can then be obtained from the
observations $i = 1, \dots, m$, say, for which $v_i \le \nu$. Analogously,
$\hat {\bm \theta}^{(B)}$ is obtained by maximizing the likelihood for the
observations $i = m + 1, \dots, n$, for which $v_i > \nu$. The LR test statistic
for the given threshold $\nu$ is then
\begin{equation} \label{eq:lr}
  \mathit{LR}(\nu) ~=~ -2 \left[
         \ell(\hat {\bm \theta}; x_1, \dots, x_n)
   ~-~ \{\ell(\hat {\bm \theta}^{(A)}; x_1, \dots, x_m)
    +    \ell(\hat {\bm \theta}^{(B)}; x_{m+1}, \dots, x_n)\}
    \right],
\end{equation}
which has an asymptotic $\chi^2$ with degrees of freedom equal to the number
of parameters in~${\bm \theta}$.

@@

For the maximum LR/Wald/LM tests, it is natural to graph the sequence of LR/Wald/LM
statistics along $V$ along with a boundary corresponding to the critical value.
Again, a boundary crossing signals a significant violation, and the peak(s) in the
sequence of statistics conveys threshold(s). Note that, due to summing over all
parameters, no specific parameter can be identified that is responsible for the
violation. Similarly,  neither component(s) nor threshold(s)
can be formally identified for the Cram\'{e}r-von Mises test. However, graphing of (transformations of) the cumulative
score process may still be valuable for gaining some insights (see,
e.g., Figure~\ref{fig:gefp_3_19}).

@@

% stopped here Feb 10 2011

@@

The figures convey information about several properties
of the tests. First, all three tests are more powerful (and in
this example significant) if only those parameters are tested
that are subject to instabilities. Conversely, if all 19~model
parameters are assessed (including those that are in fact invariant),
the power is decreased. This decrease, however, is less pronounced
for the double-max test as it is more 
sensitive to fluctuations among a small subset of parameters.

Figure~\ref{fig:lrlm} 
compares the max-LM statistic (solid line) to the max-LR
statistic (dashed line) from \eqref{eq:maxlr}, as applied to testing
\eqref{eq:h0}.  The critical values for these two
tests are identical, hence the single horizontal line.
The figure
shows that the two statistics are very similar to one another, with
both maxima at the dotted vertical line.  This
is generally to be expected, because the two tests are asymptotically
equivalent.  More generally, the 
max-LR statistic cannot be obtained from the empirical fluctuation
process, so the factor analysis model must be refitted before and after
each of the possible threshold values $\nu$.
\readme{With the modified DGP code, both max-LR and max-LM, are
nonsignificant in the example.}

@@

\fixme{I'm not really happy with the following paragraph. I think
I wouldn't call this a posthoc test. It is really a test of the same
hypothesis as before, but now using the double-max
statistic. Would it be possible to simply show this and say that
it is an advantage of the double-max test that both the threshold
and the parameter affected by the invariance can be ``seen'' in the
visualization. Then, we can still go on and talk about the power issue:
The double-max test potentially wastes a lot of power, namely
in the case where many of the parameters change at the same time. Note
that in the case where all parameters change, we wouldn't have to ask
the question ``Which parameter did change?''. Thus, luckily, the double-max test
is more powerful when we need it. But, unfortunately, we don't know that
in advance.}

The double-max test yields specific information about 
parameters violating measurement invariance.  Given
rejection of the hypothesis of measurement invariance \eqref{eq:exh0},
we can do posthoc 
tests to study specific parameters violating measurement invariance.
This involves individual examination of each parameter's cumulative
score process,
along with a procedure that controls for Type I error.  
Figure~\ref{fig:lambdas} shows the individual cumulative score processes for the
verbal factor loadings, with the horizontal lines reflecting the
critical value at $\alpha=.05$.
  The figure
shows that the third parameter (i.e., $\lambda_{31}$) crosses the
dashed line, so we would conclude a measurement 
invariance violation with respect to age for the third
verbal test.  The fact that the cumulative score process for the first
and second loadings did not achieve the critical value represents a Type II
error, which implicitly brings into question the posthoc tests' power.  We
generally address the issue of power in the simulations below.

@@

\begin{figure}
\caption{Posthoc tests of individual verbal factor loadings, based on
  the absolute maximum of each cumulative score process.  The top
  panel depicts the process for $\lambda_{11}$, the middle panel
  depicts the process for $\lambda_{21}$, and the bottom panel depicts
  the process for $\lambda_{31}$.  The solid, horizontal lines
  correspond to the critical value at $\alpha=.05$.}
\label{fig:lambdas}
\includegraphics[height=5in]{gefp_lambdas.pdf}
\end{figure}

@@

\section{Discussion}
In this paper, we have presented a new family of statistical tests for
the study of measurement invariance in psychometrics.  The tests,
based on stochastic processes, have reasonable power, can isolate
groups of individuals violating measurement invariance based on a
continuous auxiliary variable, and can
isolate specific model parameters affected by the violation.  In this
section, we consider the tests' use in practice and their 
extension to more complex scenarios.

@@

% Researchers interpreting subgroups violating MI
In addition to simultaneously testing different types of measurement
invariance, the proposed tests allow researchers to easily interpret
the nature of the invariance violation.  This is made possible through
the tests' abilities to estimate $\nu$, the threshold dividing
individuals into subgroups that violate
measurement invariance.  While a single $\nu$ was assumed in this
paper, it is also possible to define 
formal rules for estimating multiple $\nu$ parameters.  We plan to
explore these ideas in the future.

@@

One issue that was largely unaddressed in this paper involved the use
of categorical $V$ to study measurement invariance.  In this case,
groups are 
already specified in advance, and so traditional methods for fixed
subgroups (i.e., LR, Wald, and LM tests) may suffice.
Furthermore, we can also obtain an LM-type statistic from the
framework developed here. Assume the observations are divided into
$C$ categories $I_1, I_2, \ldots, I_C$. Then, the increment of
the cumulative score process $\Delta_{I_c} {\bm B}(\hat {\bm \theta})$
within in each category is just the sum of the corresponding scores.
In somewhat sloppy notation:
\begin{equation}
    \label{eq:catsum}
    \Delta_{I_c} {\bm B}(\hat {\bm \theta}) ~=~ \hat {\bm I}^{-1/2} n^{-1/2}
    \sum_{i \in I_c} {\bm s}(\hat {\bm \theta}; x_i)
%% \\
%%    X^2 & = & \sum_{c = 1}^C n_c^{-1} n^{-1} || \Delta_{I_c} {\bm B}(\hat {\bm \theta}) ||^2
\end{equation}
This results in a $C \times k$ matrix, with one entry for
each category-by-model parameter combination.  We can test
a specific model parameter for invariance by focusing on the associated
column of the $C \times k$ matrix and employing a weighted squared sum of
the entries in the column to obtain a $\chi^2$-distributed statistic with
$(C-1)$ degrees of freedom \cite{HjoKon02}.
Alternatively, to simultaneously test multiple parameters, we can sum
the $\chi^2$ statistics and degrees of freedom for the individual
parameters.

\readme{Should we also say something about ordinal variables and
variables with a lot of ties? These may also be better treated as categorical
variables if the number of potential thresholds is very low -- because
the asymptotics go in the number of thresholds.}

@@

\readme{For PATHMOX, I would just use the argument that it is not a formal
test. Whether or not it is computationally more expensive would only be
a second argument, but I would probably avoid to use it.}

@@

%\subsubsection{Other Latent Variable Models}
Aside from recursive measurement invariance detection, the proposed tests
readily extend to other popular psychometric models.  
Instead of studying measurement invariance in factor
analysis, the tests can be used to generally 
study the stability of structural equation model parameters across
observations.  For 
example, it would be possible to assess whether relationships between
latent variables are stronger for some students than for
others.  Secondly, 
the tests may be extended to study differential item functioning (DIF)
in item response models (e.g., \citeNP{KopZei10}, who focused on recursive
partitioning of Rasch models).  
Traditional DIF methods are similar to those for factor analysis in
that subgroups must be specified in advance.
%% The tests proposed here can detect subgroups automatically, however, offering a unified way of
%% studying both  measurement invariance in factor analysis and differential item
%% functioning in item response.
While these two methods have developed largely independently of one another,
some treatments \cite{Mcd99} and recent research \cite{StaChe06} have
sought to unify the methods.  Extensions of the proposed tests can
support this endeavor.
\readme{I have omitted (= commented) one sentence in the last paragraph.
I feel the paragraph may be somewhat more intelligible like this.}

@@

\readme{Instead of just giving a single URL, I prefer to give some more details
about the computational methods used. Especially in this case this makes sense,
because an OpenMx paper is forthcoming in Psychometrika.}

>>>>

To employ the LR test for assessing measurement invariance, model
parameters are estimated separately for a certain number of subgroups
of the data (with some parameters potentially restricted to be equal across
subgroups).\footnote{For ease of exposition, we describe
the case where there are no such parameter restrictions.  As shown in
the example and simulation below, it is
straightforward to extend all methods to the more general case.}
Then, the sum of maximized likelihoods 
from the subgroups are compared with the original maximized full-sample likelihood
in a $\chi^2$~test. More specifically, for the special case of two subgroups,
the alternative $H_1^*$ from (\ref{eq:h1*}) with fixed and prespecified $\nu$ is adopted
and the null hypothesis $H_0$ from (\ref{eq:h0}) reduces to ${\bm \theta}^{(A)} = {\bm \theta}^{(B)}$.
The parameter estimated $\hat {\bm \theta}^{(A)}$ can then be obtained from the
observations $i = 1, \dots, m$, say, for which $v_i \le \nu$. Analogously,
$\hat {\bm \theta}^{(B)}$ is obtained by maximizing the likelihood for the
observations $i = m + 1, \dots, n$, for which $v_i > \nu$. The LR test statistic
for the given threshold $\nu$ is then
\begin{equation} \label{eq:lr}
  \mathit{LR}(\nu) ~=~ -2 \left[
         \ell(\hat {\bm \theta}; x_1, \dots, x_n)
   ~-~ \{\ell(\hat {\bm \theta}^{(A)}; x_1, \dots, x_m)
    +    \ell(\hat {\bm \theta}^{(B)}; x_{m+1}, \dots, x_n)\}
    \right],
\end{equation}
which has an asymptotic $\chi^2$ with degrees of freedom equal to the number
of parameters in~${\bm \theta}$.

@@

For the maximum LR/Wald/LM tests, it is natural to graph the sequence of LR/Wald/LM
statistics along $V$ along with a boundary corresponding to the critical value.
Again, a boundary crossing signals a significant violation, and the peak(s) in the
sequence of statistics conveys threshold(s). Note that, due to summing over all
parameters, no specific parameter can be identified that is responsible for the
violation. Similarly,  neither component(s) nor threshold(s)
can be formally identified for the Cram\'{e}r-von Mises test. However, graphing of (transformations of) the cumulative
score process may still be valuable for gaining some insights (see,
e.g., Figure~\ref{fig:cusumex}).

@@

@@

The figures convey information about several properties
of the tests. First, all three tests are more powerful (and in
this example significant) if we test only those parameters 
that are subject to instabilities. Conversely, if all 19~model
parameters are assessed (including those that are in fact invariant),
the power is decreased. This decrease, however, is less pronounced
for the double-max test as it is more 
sensitive to fluctuations among a small subset of parameters.

Figure~\ref{fig:lrlm} 
compares the max-LM statistic (solid line) to the max-LR
statistic (dashed line) from \eqref{eq:maxlr}, as applied to testing
\eqref{eq:h0}.  The critical values for these two
tests are identical, hence the single horizontal line.
The figure
shows that the two statistics are very similar to one another, with
both maxima at the dotted vertical line.  This
is generally to be expected, because the two tests are asymptotically
equivalent.  The 
max-LR statistic cannot be obtained from the empirical fluctuation
process, however, so the factor analysis model must be refitted before
and after each of the possible threshold values $\nu$.

@@

\readme{Achim previous comment: I'm not really happy with the following
  paragraph. I think 
I wouldn't call this a posthoc test. It is really a test of the same
hypothesis as before, but now using the double-max
statistic. \\ \ \\
Ed response: I am fine with not calling it a posthoc test.  However,
it could sometimes be advantageous to first test the hypothesis using the CvM
statistic, then, if CvM rejects the hypothesis, employ the double-max
statistic to say something about individual model parameters.  This
would be useless if all parameters change, but, for measurement
invariance problems, one often expects many model parameters to be stable.}

As mentioned previously, the double-max test is advantageous because
it yields information about individual 
parameters violating measurement invariance.  
That is, it allows us to examine whether or not individual parameters' 
cumulative score processes lead one to reject the hypothesis of
measurement invariance.
Figure~\ref{fig:lambdas} shows the individual cumulative score processes for the
verbal factor loadings, with the horizontal lines reflecting the
critical value at $\alpha=.05$.
  The figure
shows that the third parameter (i.e., $\lambda_{31}$) crosses the
dashed line, so we would conclude a measurement 
invariance violation with respect to age for the third
verbal test.  The fact that the cumulative score process for the first
and second loadings did not achieve the critical value represents a Type II
error, which implicitly brings into question the posthoc tests' power.
This is an example of a situation where the double-max test ``wastes''
power, as it cannot make use of the fact that multiple model
parameters change simultaneously.  We
generally address the issue of power in the simulations below.

@@

\begin{figure}
\caption{Cumulative score processes for each verbal factor
  loadings, with critical values stemming from the double-max test. 
  The solid, horizontal lines
  correspond to the critical value at $\alpha=.05$.}
\label{fig:lambdas}
\includegraphics[height=5in]{gefp_lambdas.pdf}
\end{figure}

@@

\section{Discussion}
In this paper, we have presented a new family of statistical tests for
the study of measurement invariance in psychometrics.  The tests,
based on stochastic processes, have reasonable power, can isolate
subgroups of individuals violating measurement invariance based on a
continuous auxiliary variable, and can
isolate specific model parameters affected by the violation.  In this
section, we consider the tests' use in practice and their 
extension to more complex scenarios.

@@

% Researchers interpreting subgroups violating MI
In addition to simultaneously testing different types of measurement
invariance, the proposed tests allow researchers to easily interpret
the nature of the invariance violation.  This is made possible through
the tests' abilities to estimate $\nu$, the threshold dividing
individuals into subgroups that violate
measurement invariance.  While a single $\nu$ was assumed in this
paper, it is also possible to define 
formal rules for estimating multiple $\nu$ parameters.  We plan to
explore these issues in the future.

@@

One issue that was largely unaddressed in this paper involved the use
of categorical $V$ to study measurement invariance.  In this case,
groups are 
already specified in advance, and so traditional methods for fixed
subgroups (i.e., LR, Wald, and LM tests) may suffice.
Furthermore, we can also obtain an LM-type statistic from the
framework developed here. Assume the observations are divided into
$C$ categories $I_1, I_2, \ldots, I_C$. Then, the increment of
the cumulative score process $\Delta_{I_c} {\bm B}(\hat {\bm \theta})$
within in each category is just the sum of the corresponding scores.
In somewhat sloppy notation:
\begin{equation}
    \label{eq:catsum}
    \Delta_{I_c} {\bm B}(\hat {\bm \theta}) ~=~ \hat {\bm I}^{-1/2} n^{-1/2}
    \sum_{i \in I_c} {\bm s}(\hat {\bm \theta}; x_i)
%% \\
%%    X^2 & = & \sum_{c = 1}^C n_c^{-1} n^{-1} || \Delta_{I_c} {\bm B}(\hat {\bm \theta}) ||^2
\end{equation}
This results in a $C \times k$ matrix, with one entry for
each category-by-model parameter combination.  We can test
a specific model parameter for invariance by focusing on the associated
column of the $C \times k$ matrix and employing a weighted squared sum of
the entries in the column to obtain a $\chi^2$-distributed statistic with
$(C-1)$ degrees of freedom \cite{HjoKon02}.
Alternatively, to simultaneously test multiple parameters, we can sum
the $\chi^2$ statistics and degrees of freedom for the individual
parameters.  In addition to categorical $V$, this framework may also
be useful for continuous $V$ with many ties and ordinal $V$.  
The number of potential thresholds may be very low in these
situations, which impacts the extent to which asymptotic results hold
for the main test statistics described in this paper.

@@

@@

%\subsubsection{Other Latent Variable Models}
The proposed tests also 
readily extend to other popular psychometric models.  
Instead of studying measurement invariance in factor
analysis, the tests can be used to generally 
study the stability of structural equation model parameters across
observations.  For 
example, it would be possible to assess whether relationships between
latent variables are stronger for some individuals than for
others.  Secondly, 
the tests may be extended to study differential item functioning (DIF)
in item response models (e.g., \citeNP{KopZei10}, who focused on recursive
partitioning of Rasch models).  
Traditional DIF methods are similar to those for factor analysis in
that subgroups must be specified in advance.
%% The tests proposed here can detect subgroups automatically, however, offering a unified way of
%% studying both  measurement invariance in factor analysis and differential item
%% functioning in item response.
While factor-analytic measurement invariance methods and DIF methods
have developed 
largely independently of one another, 
some treatments \cite{Mcd99} and recent research \cite{StaChe06} have
sought to unify the methods.  Extensions of the proposed tests can
support this endeavor.

@@
